---
title: "STAT571 HW1 Group 18 - Case 3"
output: html_notebook
---


Case 3: MLB

## Setup & Imports
```{r}
knitr::opts_chunk$set(echo = FALSE, results = "hide", fig.width=8, fig.height=4) # do formatting for eventual knitted output
options(scipen = 0, digits = 3)  # controls base R output
if(!require('pacman')) {install.packages('pacman')} # if you don't have pacman, install it
pacman::p_load(ISLR, readxl, tidyverse, magrittr, dplyr, ggplot2) # load necessary packages
```


## Read in Data
```{r}
baseball <- read.csv(file = 'data/baseball.csv') # read in the long-format csv as a data.frame
head(baseball, 5) # display the first 5 rows
```

## 4.1 EDA: Relationship between payroll changes and performance

First, we create a new variable called diff_lag, which describes the log difference between a year's payroll and the previous years' payroll.

In this setup, it is more appropriate to work with log-transformed data because it is more informative about *relative* differences than the raw amount of payroll increase. In other words, the same amount of raw change in payroll would represent a more significant increase to a team earning less, than a team earning more, so we can take the differences of the log values to address this. 
```{r}
baseball$diff_log <-  log(baseball$payroll) - log(lag(baseball$payroll)) #here we create the new variable of differences of log values
baseball$diff_win_pct <-  baseball$win_pct - lag(baseball$win_pct) # I'm also going to create a variable that describes the change in win_pct, which will help in 4.2 

# now, we create a new table with variables of interest: team, year, diff_log, and win_pct
baseball_variables <- baseball %>% 
  select(team, year, diff_log, payroll, win_pct, diff_win_pct) %>%
  na.omit # Because the first year in our data will not have a computable difference from the previous year, it will produce an NA in the diff_log column. This particular row will not contain information relevant to our analysis, so we can simply omit the rows that have NAs. 


head(baseball_variables, 5) # let's take a look to make sure we've got what we need
```

## 4.2
```{r}

# for convenience, create a table with just the years 2010-2014
baseball_variables_2010_2014 <- baseball_variables %>%
  subset(year >= 2010 & year <= 2014) # we're only interested in the years between 2010 and 2014 inclusive

# sum up each team's change in payroll over the years. Here I also do the same for the chang ein win percentage, as this will aid our subsequent comparison. 
baseball_variables_2010_2014 <- aggregate(cbind(diff_log,diff_win_pct ) ~ team, data = baseball_variables_2010_2014, FUN = sum)

# now, let's see which 5 teams had the largest increase in payroll between 2010 and 2014
baseball_variables_2010_2014 %>%
  arrange(desc(diff_log)) %>% 
  head(5)

# which 5 teams had the largest increase in win percentage between 2010 and 2014
baseball_variables_2010_2014 %>%
  arrange(desc(diff_win_pct)) %>% 
  head(5)

```
As we can see in the procedure above, the top 5 teams with the highest *payroll* increase between 2010 and 2014 are:
1. Los Angeles Dodgers
2. Washington Nationals
3. San Diego Padres
4. Texas Rangers
5. San Fransisco Giants

And the top 5 teams with the highest win percentage increase between 2010 and 2014 are: 
1. Washington Nationals
2. Baltimore Orioles
3. Pittsburgh Pirates
4. Kansas City Royals
5. Cleveland Indians

## 4.3
```{r}

cor(baseball_variables$diff_log, baseball_variables$diff_win_pct, method = 'pearson') # check the correlation between the two variables of interest (diff_log and change in win_pct)

# fit a linear model to see if diff_log predicts diff_win_pct
baseball_model <- lm(diff_win_pct ~ diff_log, data = baseball_variables)
summary(baseball_model) # lm output

# ok, let's plot the lm 
ggplot(baseball_variables, aes(x = diff_log, y = diff_win_pct)) + geom_point() + stat_smooth(method = "lm", col = "red")

```
It appears based on this analysis that while there does seem to be a statistically significant association between log increase in pay and increase in performance (p = 0.043), the effect size is small (Adj. R-Squared = 0.00612). Accordingly, my interpretation is that while there is in fact evidence that an increase in payroll may lead to an increase in performance, it is likely that change in payroll alone does not adequately explain performance. 

## 4.4
```{r}
baseball_model_2 <- lm(diff_win_pct ~ payroll, data = baseball_variables)
summary(baseball_model_2) # lm output

baseball_model_complete <- lm(diff_win_pct ~ diff_log + payroll, data = baseball_variables)
summary(baseball_model_complete) # lm output

```

It appears that between change in log payroll, and payroll as such, the change in log payroll is more predictive of change in performance. 
It should be noted that, if we create a linear model that combines both payroll as such, and change in log payroll, it better predicts change in performance than either variable alone. 




