---
title: "Modern Data Mining - HW 5"
author:
- Diego G. Dávila
- Margaret Gardner
- Joelle Bagautdinova
date: 'Due: 11:59Pm,  4/10, 2022'
output:
  html_document:
    code_folding: show
    highlight: haddock
    theme: lumen
    toc: yes
    toc_depth: 4
    toc_float: yes
  pdf_document:
    toc: yes
    toc_depth: '4'
  word_document:
    toc: yes
    toc_depth: '4'
urlcolor: blue  
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(fig.height=4, fig.width=6, warning = F)
if (!require("pacman")) install.packages("pacman")
pacman::p_load(randomForest, tree, ISLR, rpart, rattle, pROC, partykit, ggplot2, glmnet, leaps, dplyr, keras, neuralnet, imager, ranger, skimr, car, data.table)
```




# Overview

For the purpose of predictions, a model free approach could be beneficial. A binary decision tree is the simplest, still interpretable and often provides insightful information between predictors and responses. To improve the predictive power we would like to aggregate many equations, especially uncorrelated ones. One clever way to have many free samples is to take bootstrap samples. For each bootstrap sample we  build a random tree by taking a randomly chosen number of variables to be split at each node. We then take average of all the random bootstrap trees to have our final prediction equation. This is RandomForest. 

Ensemble method can be applied broadly: simply take average or weighted average of many different equations. This may beat any single equation in your hand.


All the methods covered can handle both continuous responses as well as categorical response with multiple levels (not limited to binary response.)


## Objectives


- Understand trees
    + single tree/displaying/pruning a tree
    + RandomForest
    + Ensemble idea

- R functions/Packages
    + `tree`, `RandomForest`, `ranger`
    
- Json data format

- text mining
    + bag of words
  

Data needed:

+ `IQ.Full.csv`
+ `yelp_review_20k.json`

# Problem 0: Lectures

Please study all three lectures. Understand the main elements in each lecture and be able to run and compile the lectures

+ textmining
+ trees
+ boosting




# Problem 1: IQ and successes

## Background: Measurement of Intelligence 

Case Study:  how intelligence relates to one's future successes?

**Data needed: `IQ.Full.csv`**

ASVAB (Armed Services Vocational Aptitude Battery) tests have been used as a screening test for those who want to join the army or other jobs. 

Our data set IQ.csv is a subset of individuals from the 1979 National Longitudinal Study of 
Youth (NLSY79) survey who were re-interviewed in 2006. Information about family, personal demographic such as gender, race and education level, plus a set of ASVAB (Armed Services Vocational Aptitude Battery) test scores are available. It is STILL used as a screening test for those who want to join the army! ASVAB scores were 1981 and income was 2005. 

**Our goals:** 

+ Is IQ related to one's successes measured by Income?
+ Is there evidence to show that Females are under-paid?
+ What are the best possible prediction models to predict future income? 


**The ASVAB has the following components:**

+ Science, Arith (Arithmetic reasoning), Word (Word knowledge), Parag (Paragraph comprehension), Numer (Numerical operation), Coding (Coding speed), Auto (Automative and Shop information), Math (Math knowledge), Mechanic (Mechanic Comprehension) and Elec (Electronic information).
+ AFQT (Armed Forces Qualifying Test) is a combination of Word, Parag, Math and Arith.
+ Note: Service Branch requirement: Army 31, Navy 35, Marines 31, Air Force 36, and Coast Guard 45,(out of 100 which is the max!) 

**The detailed variable definitions:**

Personal Demographic Variables: 

 * Race: 1 = Hispanic, 2 = Black, 3 = Not Hispanic or Black
 * Gender: a factor with levels "female" and "male"
 * Educ: years of education completed by 2006
 
Household Environment: 
 
* Imagazine: a variable taking on the value 1 if anyone in the respondent’s household regularly read
	magazines in 1979, otherwise 0
* Inewspaper: a variable taking on the value 1 if anyone in the respondent’s household regularly read
	newspapers in 1979, otherwise 0
* Ilibrary: a variable taking on the value 1 if anyone in the respondent’s household had a library card
	in 1979, otherwise 0
* MotherEd: mother’s years of education
* FatherEd: father’s years of education

Variables Related to ASVAB test Scores in 1981 (Proxy of IQ's)

* AFQT: percentile score on the AFQT intelligence test in 1981 
* Coding: score on the Coding Speed test in 1981
* Auto: score on the Automotive and Shop test in 1981
* Mechanic: score on the Mechanic test in 1981
* Elec: score on the Electronics Information test in 1981

* Science: score on the General Science test in 1981
* Math: score on the Math test in 1981
* Arith: score on the Arithmetic Reasoning test in 1981
* Word: score on the Word Knowledge Test in 1981
* Parag: score on the Paragraph Comprehension test in 1981
* Numer: score on the Numerical Operations test in 1981

Variable Related to Life Success in 2006

* Income2005: total annual income from wages and salary in 2005. We will use a natural log transformation over the income.


**Note: All the Esteem scores shouldn't be used as predictors to predict income**

## 1. EDA: Some cleaning work is needed to organize the data. 

+ The first variable is the label for each person. Take that out.
+ Set categorical variables as factors. 
+ Make log transformation for Income and take the original Income out
+ Take the last person out of the dataset and label it as **Michelle**. 
+ When needed, split data to three portions: training, testing and validation (70%/20%/10%)
  - training data: get a fit
  - testing data: find the best tuning parameters/best models
  - validation data: only used in your final model to report the accuracy. 


```{r}
# load data
df_full <- read.csv("data/IQ.Full.csv")

# data cleaning
df_full <- df_full %>% 
  # seet categorical variables as factors
  mutate(Race = as.factor(Race), 
         Gender = as.factor(Gender),
         Imagazine = as.factor(Imagazine),
         Inewspaper = as.factor(Inewspaper),
         Ilibrary = as.factor(Ilibrary),
         log_income = log(Income2005)
  )  %>%
  # remove ID and original income variables
  select(-c(Subject, Income2005), -contains(c("Esteem")))

# label last row as Michelle
Michelle <- tail(df_full, n = 1)

# remove the last row
df <- df_full %>%
  filter(row_number() <= n() - 1)

skim(df)

# split data into training, testing and validation
N <- dim(df)[1]
n1 <- floor(.7*N) # training = 70%
n2 <- floor(.2*N) # testing = 20%

set.seed(10)

# Split data to three portions of 70%/20%/10% of data size N
idx_train <- sample(N, n1)
idx_no_train <- (which(! seq(1:N) %in% idx_train))
idx_test <- sample(idx_no_train, n2)
idx_val <- which(! idx_no_train %in% idx_test)
data.train <- as.data.frame(df[idx_train,])
data.test <- as.data.frame(df[idx_test,])
data.val <- as.data.frame(df[idx_val,])

```


## 2. Factors affect Income

We only use linear models to answer the questions below.

i. To summarize ASVAB test scores, create PC1 and PC2 of 10 scores of ASVAB tests and label them as
ASVAB_PC1 and ASVAB_PC2. Give a quick interpretation of each ASVAB_PC1 and ASVAB_PC2 in terms of the original 10 tests. 

```{r}

# create ASVAB subset
ASVAB <- df[, 10:19]

# apply PCA with scaling and centering
ASVAB_pca <- prcomp(ASVAB, scale. = TRUE, center = TRUE)

# display the loadings
ASVAB_pca$rotation

# add PC1 and PC2 to the df
df$ASVAB_PC1 <- ASVAB_pca$x[,"PC1"]
df$ASVAB_PC2 <- ASVAB_pca$x[,"PC2"]


# redo the same for the full dataset and reselect Michelle's row (having PC1 and PC2 for Michelle is necessary for the tree section below)
ASVAB <- df_full[, 10:19]
ASVAB_pca <- prcomp(ASVAB, scale. = TRUE, center = TRUE)
df_full$ASVAB_PC1 <- ASVAB_pca$x[,"PC1"]
df_full$ASVAB_PC2 <- ASVAB_pca$x[,"PC2"]

# label last row as Michelle once again (this time with the PCs)
Michelle <- tail(df_full, n = 1)


```


**Interpretation:** 

* ASVAB_PC1 appears to be a weighted sum of all 10 items and can thus be interpreted as a general ASVAB score. 
* ASVAB_PC2 appears to capture the difference between performance in more abstract reasoning (Arith, Word, Parag, Numer, Coding, Math) vs more applied tasks (Science, Auto, Mechanic, Elec). 

ii. Is there any evidence showing ASVAB test scores in terms of ASVAB_PC1 and ASVAB_PC2, might affect the Income?  Show your work here. You may control a few other variables, including gender. 

```{r}

# let's start with a simple linear model
fit1 <- lm(log_income ~ ASVAB_PC1 + ASVAB_PC2, data = df)
summary(fit1)

fit2 <- lm(log_income ~ ASVAB_PC1 + ASVAB_PC2 + Gender, data = df)
summary(fit2)

fit3 <- lm(log_income ~ ASVAB_PC1 + ASVAB_PC2 + Gender + Race, data = df)
Anova(fit3)

fit4 <- lm(log_income ~ ASVAB_PC1 + ASVAB_PC2 + Gender + Educ, data = df)
summary(fit4)

fit5 <- lm(log_income ~ ASVAB_PC1 + ASVAB_PC2 + Gender + Educ + MotherEd + FatherEd, data = df)
summary(fit5)
```

**Answer:** Based on the linear models above, the following observations can be made: 

* ASVAB_PC1 and ASVAB_PC2 both have a significant effect on log_outcome, where having a higher score on either PC is associated with higher income. 
* Gender has a significant effect, with males earning a higher income compared to females. 
* Years of education have a significant effect on log_income, where higher years of education for the individual or their father is associated with higher log_income. 
* Race or years of the mother's or father's education do not significantly impact log_income.

iii. Is there any evidence to show that there is gender bias against either male or female in terms of income in the above model? 

```{r}

# display results of model fit including Gender
Anova(fit2)
summary(fit2)

# testing the difference between the reduced model (without Gender) and full model (including Gender)
anova(fit1, fit2)
```

Based on the models above, we note that Gender does have a significant effect on log_income. Specifically, males tend to earn 10.097 + 0.09 more log_income compared to females (reference level).  

We next build a few models for the purpose of prediction using all the information available. From now on you may use the three data sets setting (training/testing/validation) when it is appropriate. 

## 3. Trees

i. fit1: tree(Income ~ Educ + Gender, data.train) with default set up 

    a) Display the tree
  
    ```{r}
# fit the tree
fit1 <- tree(log_income ~ Educ + Gender, data.train)

# view numerical output
summary(fit1)

# plot the tree
plot(fit1)
title(main="log_income ~ Educ + Gender")
text(fit1)

```

    
    b) How many end nodes? Briefly explain how the estimation is obtained in each end nodes and describe the prediction equation
    
**Answer:** The tree has 4 end nodes, meaning that Gender and Education are partitioned into 4 boxes. The end node estimations correspond to the sample mean log_income for every end node. The prediction equation for this model is: 

    c) Does it show interaction effect of Gender and Educ over Income?
    
**Answer"** Yes, the tree shows an interaction between gender and income, where being male with more than 15.5 years of education corresponds to the highest log_income, and conversely being female with less than 15.5 years of education is associated with the lowest log_income. 

    d) Predict Michelle's income

```{r}

fit.Michelle <- predict(fit1, Michelle)
fit.Michelle # the fitted salary in log scale is 6.196343 (may not be the same each time we run the rf, why?)

Michelle_pred_income <- exp(fit.Michelle)

```

**Answer"** Michelle's predicted log_income is `r fit.Michelle`, which corresponds to $ `r Mihcelle_pred_income`. 

ii. fit2: fit2 <- rpart(Income2005 ~., data.train, minsplit=20, cp=.009)

    a) Display the tree using plot(as.party(fit2), main="Final Tree with Rpart") 
    
```{r}

# fit tree using rpart
fit2 <- rpart(log_income ~., data.train, minsplit = 20, cp = .009)

# plot the rpart tree
plot(as.party(fit2), main = "Final Tree with Rpart") 

```

    b) A brief summary of the fit2

This full tree inclusing all predictors clearly illustrates the effect of gender on log_outcome, where being female leads to a lower income. The next splits involve several ASVAB scores (Math, Arith), years of education and Esteem10. Overall, males with a high score in Arithmetic and at least 15.5 years of education are predicted to have the highest income. On the other hand, females with a Math score under 10.5 are predicted to earn the lowest income. 

    c) Compare testing errors between fit1 and fit2. Is the training error from fit2 always less than that from fit1? Is the testing error from fit2 always smaller than that from fit1? 
    
```{r}

# training error
fit1.rss <- summary(fit1)$dev # fit1 RSS
fit2.rss <- sum(residuals(fit2)^2) # fit2 RSS (fit2 has a different output as rpart was used)

# testing error
test.error.fit1 <- sum((predict(fit1, data.test) - data.test$log_income)^2)
test.error.fit2 <- sum((predict(fit2, data.test) - data.test$log_income)^2)
```

**Answer:** The training error of fit2 (`r fit2.rss`) is smaller than the training error of fit1 (`r fit1.rss`). However, the testing error of fit2 (`r test.error.fit2`) is slightly larger than the testing error of fit1 (`r test.error.fit1`). While the training error always should be smaller for trees that include more variables, this does not necessarily hold true for the testing error, as this corresponds to the model fit on unseen data.  **CHECK THIS**

    d) You may prune the fit2 to get a tree with small testing error. 
    
```{r}

### testing a few pruning options ###

# first print the cps of the rpart object
# Note: cp = complexity parameter. Any split that does not decrease the overall lack of fit by a factor of cp is not attempted.
printcp(fit2)
plotcp(fit2) # plot the cps

# find the option with the smallest error (in xerror)
fit2$cptable[which.min(fit2$cptable[,"xerror"]),"CP"] # looks like 0.009 was already the optimal cp!

# let's try several of these cp options on the test data to find their testing error 


# cp = 0.07
fit2.p1 <- prune(fit2, cp = 0.07)  # prune fit2 using prune() from the rpart package (as the tree has been generated with rpart)
print(paste("cp = 0.07: ", round(sum((predict(fit2.p1, data.test) - data.test$log_income)^2), 3))) # testing error 

# cp = 0.033
fit2.p2 <- prune(fit2, cp = 0.033)  # prune fit2 using prune() from the rpart package (as the tree has been generated with rpart)
print(paste("cp = 0.033: ", round(sum((predict(fit2.p2, data.test) - data.test$log_income)^2), 3))) # testing error 

# cp = 0.02
fit2.p3 <- prune(fit2, cp = 0.02)  # prune fit2 using prune() from the rpart package (as the tree has been generated with rpart)
print(paste("cp = 0.02: ", round(sum((predict(fit2.p3, data.test) - data.test$log_income)^2), 3))) # testing error 

# cp = 0.014
fit2.p4 <- prune(fit2, cp = 0.014)  # prune fit2 using prune() from the rpart package (as the tree has been generated with rpart)
print(paste("cp = 0.014: ", round(sum((predict(fit2.p4, data.test) - data.test$log_income)^2), 3))) # testing error 

# cp = 0.011
fit2.p5 <- prune(fit2, cp = 0.011)  # prune fit2 using prune() from the rpart package (as the tree has been generated with rpart)
print(paste("cp = 0.011: ", round(sum((predict(fit2.p5, data.test) - data.test$log_income)^2), 3))) # testing error 

# cp = 0.0091
fit2.p6 <- prune(fit2, cp = 0.0091)  # prune fit2 using prune() from the rpart package (as the tree has been generated with rpart)
print(paste("cp = 0.0091: ", round(sum((predict(fit2.p6, data.test) - data.test$log_income)^2), 3))) # testing error 

```

  
**Answer:** As it turns out, while cp = 0.009 has the smallest training error, cp = 0.011 has the smallest testing error. Thus, let's build the pruned tree using this criterion: 

```{r}
# summary(fit2.pruned)
data.table(fit2.p5$frame)

#plot the best subtrees
par(mfrow=c(1, 2))
plot(fit2) # main tree
title(main = "Main tree")
text(fit2) # main tree
plot(fit2.p5)
title(main="Best pruned tree (cp = 0.011)")
text(fit2.p5) # pruned tree

```

  
iii. fit3: bag two trees

    a) Take 2 bootstrap training samples and build two trees using the 
    rpart(Income2005 ~., data.train.b, minsplit=20, cp=.009). Display both trees.
    
```{r}

# bootstrap tree 1 
par(mfrow=c(1, 2))
n=nrow(df)
set.seed(1)  
index1 <- sample(n, n, replace = TRUE)
df_b1 <- df[index1, ]  #  first bootstrap sample
boot.1 <- rpart(log_income ~., df_b1, minsplit = 20, cp = .009) 
plot(boot.1)
title(main = "First bootstrap tree")
text(boot.1, pretty = 0)

# bootstrap tree 2 
set.seed(2) # will be a different sample
index1 <- sample(n, n, replace = TRUE)
df_b2 <- df[index1, ]  # second bootstrap sample
boot.2 <- rpart(log_income ~., df_b2, minsplit = 20, cp = .009) 
plot(boot.2)
title(main = "Second bootstrap tree")
text(boot.2, pretty = 0)
```


    b) Explain how to get fitted values for Michelle by bagging the two trees obtained above. Do not use the predict(). 
    
```{r}

# Normally, we would bag the two trees by averaging the two fitted equations and predict() the response for Michelle as follows
# fit.bag.predict <- (predict(boot.1, Michelle) + predict(boot.2, Michelle))/2 #bagging
# data.frame(fitted=fit.bag.predict,  obsy = Michelle["log_income"])

# Alternatively, bagging can also be done using a randomforest, by specifying that the number of trees = 2 and by using ALL 20 predictors (which is the equivalent of bagging and removes the randomness)
set.seed(1)
fit3 <- randomForest(log_income ~., data.train, mtry = 20, ntree = 2)

# predict Michelle based on the 2 bagged trees
fit.Michelle.2 <- predict(fit3, Michelle)
data.frame(fitted=fit.Michelle.2,  obsy = Michelle["log_income"])

# could potentially get Michelle by looking at the OOB for her - but that only works if she was included in the dataset... And they ask us to exclude her in the beginning... Don't have any other ideas on how to get her fitted values without predict...
# Add Michelle back into the data.train sample
set.seed(1)
data.train.Michelle <- rbind(data.train, Michelle)
fit3.wMichelle <- randomForest(log_income ~., data.train.Michelle, mtry = 20, ntree = 2)
fit3.wMichelle$predicted[1809]

```


**Answer:** The `randomForest()` function can be used as an alternative to get fitted values for Michelle based on two bagged trees above without using the predict function. More specifically, to do bagging of 2 trees, `ntree` has to be set to 2 and `mtry` has to be set to all included predictors (20 here), as this will lead to bagging instead of random tree building. Next, to get Michelle's fitted value without using predict(), we can add Michelle's data to the data.train sample and rerun the random forest. Her fitted log_income can then be extracted by selecting the last (newly added) row of `fit3.wMichelle$predicted[1809]`.
**Not sure how to get Michelle's fitted values without predict?? Do they just mean no use of predict for the bagging part?**

    c) What is the testing error for the bagged tree. Is it guaranteed that the testing error by bagging the two tree always smaller that either single tree? 
    
```{r}
# getting bagged tree testing error
predicted_oob <- fit3$predicted
mse_oob <- mean((data.test$log_income - predicted_oob)^2, na.rm = TRUE) # there are some NAs in the predicted values

print(paste("fit3 testing error: ", round(mse_oob, 3)))
```

**Answer:** Given that bagging involves averaging both trees, this necessarily reduces the testing error of the bagged tree compared to the original trees it is composed of.  **CHECK THIS. If not always true, can say: It is very likely that the testing error of the bagged tree will be smaller **

    
iv. fit4: Build a best possible RandomForest

    a) Show the process how you tune mtry and number of trees. Give a very high level explanation how fit4 is built.
    b) Compare the oob errors form fit4 to the testing errors using your testing data. Are you convinced that oob errors estimate testing error reasonably well.
    c) What is the predicted value for Michelle?

    
v. Now you have built so many predicted models (fit1 through fit4 in this section). What about build a fit5 which bags fit1 through fit4. Does fit5 have the smallest testing error?

vi.  Summarize the results and nail down one best possible final model you will recommend to predict income. Explain briefly why this is the best choice. Finally for the first time evaluate the prediction error using the validating data set. 

vii. Use your final model to predict Michelle's income. 


    
# Problem 2: Yelp challenge 2019

**Note:** This problem is rather involved. It covers essentially all the main materials we have done so far in this semester. It could be thought as a guideline for your final project if you want when appropriate. 

Yelp has made their data available to public and launched Yelp challenge. [More information](https://www.yelp.com/dataset/). It is unlikely we will win the $5,000 prize posted but we get to use their data for free. We have done a detailed analysis in our lecture. This exercise is designed for you to get hands on the whole process. 

For this case study, we downloaded the [data](https://www.yelp.com/dataset/download) and took a 20k subset from **review.json**. *json* is another format for data. It is flexible and commonly-used for websites. Each item/subject/sample is contained in a brace *{}*. Data is stored as **key-value** pairs inside the brace. *Key* is the counterpart of column name in *csv* and *value* is the content/data. Both *key* and *value* are quoted. Each pair is separated by a comma. The following is an example of one item/subject/sample.

```{json}
{
  "key1": "value1",
  "key2": "value2"
}
```


**Data needed:** yelp_review_20k.json available in Canvas.

**yelp_review_20k.json** contains full review text data including the user_id that wrote the review and the business_id the review is written for. Here's an example of one review.

```{json}
{
    // string, 22 character unique review id
    "review_id": "zdSx_SD6obEhz9VrW9uAWA",

    // string, 22 character unique user id, maps to the user in user.json
    "user_id": "Ha3iJu77CxlrFm-vQRs_8g",

    // string, 22 character business id, maps to business in business.json
    "business_id": "tnhfDv5Il8EaGSXZGiuQGg",

    // integer, star rating
    "stars": 4,

    // string, date formatted YYYY-MM-DD
    "date": "2016-03-09",

    // string, the review itself
    "text": "Great place to hang out after work: the prices are decent, and the ambience is fun. It's a bit loud, but very lively. The staff is friendly, and the food is good. They have a good selection of drinks.",

    // integer, number of useful votes received
    "useful": 0,

    // integer, number of funny votes received
    "funny": 0,

    // integer, number of cool votes received
    "cool": 0
}
```

## Goal of the study

The goals are 

1) Try to identify important words associated with positive ratings and negative ratings. Collectively we have a sentiment analysis.  

2) To predict ratings using different methods. 

## 1. JSON data and preprocessing data

i. Load *json* data

The *json* data provided is formatted as newline delimited JSON (ndjson). It is relatively new and useful for streaming.
```{json}
{
  "key1": "value1",
  "key2": "value2"
}
{
  "key1": "value1",
  "key2": "value2"
}
```

The traditional JSON format is as follows.
```{json}
[{
  "key1": "value1",
  "key2": "value2"
},
{
  "key1": "value1",
  "key2": "value2"
}]
```


We use `stream_in()` in the `jsonlite` package to load the JSON data (of ndjson format) as `data.frame`. (For the traditional JSON file, use `fromJSON()` function.)

```{r}
pacman::p_load(jsonlite)
yelp_data <- jsonlite::stream_in(file("data/yelp_review_20k.json"), verbose = F)
str(yelp_data)  

# different JSON format
# tmp_json <- toJSON(yelp_data[1:10,])
# fromJSON(tmp_json)
```

**Write a brief summary about the data:**

a) Which time period were the reviews collected in this data?

b) Are ratings (with 5 levels) related to month of the year or days of the week? Only address this through EDA please. 


ii. Document term matrix (dtm)
 
 Extract document term matrix for texts to keep words appearing at least .5% of the time among all 20000 documents. Go through the similar process of cleansing as we did in the lecture. 

a) Briefly explain what does this matrix record? What is the cell number at row 100 and column 405? What does it represent?

b) What is the sparsity of the dtm obtained here? What does that mean?

iii. Set the stars as a two category response variable called rating to be “1” = 5,4 and “0”= 1,2,3. Combine the variable rating with the dtm as a data frame called data2. 

## Analysis

Get a training data with 13000 reviews and the 5000 reserved as the testing data. Keep the rest (2000) as our validation data set. 

## 2. LASSO

i. Use the training data to get Lasso fit. Choose lambda.1se. Keep the result here.

ii. Feed the output from Lasso above, get a logistic regression. 
	
a) Pull out all the positive coefficients and the corresponding words. Rank the coefficients in a decreasing order. Report the leading 2 words and the coefficients. Describe briefly the interpretation for those two coefficients. 

b) Make a word cloud with the top 100 positive words according to their coefficients. Interpret the cloud briefly.

c) Repeat i) and ii) for the bag of negative words.

d) Summarize the findings. 

iii. Using majority votes find the testing errors
	i) From Lasso fit in 3)
	ii) From logistic regression in 4)
	iii) Which one is smaller?

## 3. Random Forest  

i. Briefly summarize the method of Random Forest

ii. Now train the data using the training data set by RF. Get the testing error of majority vote. Also explain how you tune the tuning parameters (`mtry` and `ntree`). 

## 4. Boosting - REMOVED

SECTION REMOVED


## 5.  PCA first

i. Perform PCA (better to do sparse PCA) for the input matrix first. Decide how many PC's you may want to take and why.

ii. Pick up one of your favorate method above and build the predictive model with PC's. Say you use RandomForest.

iii. What is the testing error? Is this testing error better than that obtained using the original x's? 

## 6. Ensemble model

i. Take average of some of the  models built above (also try all of them) and this gives us the fifth model. Report it's testing error. (Do you have more models to be bagged, try it.)


## 7. Final model

Which classifier(s) seem to produce the least testing error? Are you surprised? Report the final model and accompany the validation error. Once again this is THE only time you use the validation data set.  For the purpose of prediction, comment on how would you predict a rating if you are given a review (not a tm output) using our final model? 











